
This notebook will be used to perform differential expression analysis on the count data generated from the experiments from the study GSE272548.

```{r}
rm(list=ls())

suppressMessages({
    library(readr)              # v2.1.5
    library(DESeq2)             # v1.42.1
    library(dplyr)              # v1.1.4
    library(gtools)             # v3.9.5
    library(BiocParallel)       # v1.36.0
    library(tidyverse)          # v2.0.0
    library(qs)                 # v0.27.3
})                              # R v4.3.3 

register(MulticoreParam(workers = 40))  
```




# METADATA 

```{r}
metadata <- suppressMessages(read_csv("../input/metadata/File_S5_httr_mcf7_screen_qc_metrics.csv"))
metadata <- metadata[metadata$qc_flag == "OK", ]

exclude_stypes <- c("QC sample", "reference chemical", "untreated control")  
metadata <- metadata[ ! metadata$stype %in% exclude_stypes, ]

keep_stypes <- c("test sample", "vehicle control")
metadata <- metadata[metadata$stype %in% keep_stypes, ]

manifest <- suppressMessages(read_csv("../input/metadata/hWTv1_mcf7_probes.csv"))
```




# COUNT DATA 

```{r}
count_files <- list.files( "../input/count_data", pattern = "count_data.csv", full.names = TRUE)
```




# DE ANALYSIS

```{r}
all_plategroups <- mixedsort(unique(metadata$pg_id))
all_results <- list()

for (pg in all_plategroups) {
    cat("Processing plate group:", pg, "\n")
    
    relevant_metadata <- metadata[metadata$pg_id == pg, ]
    
    count_file_pattern <- paste0("pg", pg, "_count")
    file_index <- grepl(count_file_pattern, count_files)
    if (sum(file_index) == 0) {
        stop("No count files found for plate group ", pg)
    }
    
    relevant_count_data <- as.data.frame(
        suppressMessages(read_csv(count_files[file_index]))
    )
    
    rownames(relevant_count_data) <- relevant_count_data[[1]]
    relevant_count_data <- as.matrix(relevant_count_data[,-1])
    
    common_samples <- intersect(colnames(relevant_count_data), relevant_metadata$sample_id)
    if (length(common_samples) == 0) {
        stop("No common samples found for plate group ", pg)
    }
    
    count_matrix <- relevant_count_data[, common_samples, drop = FALSE]
    relevant_metadata <- relevant_metadata[match(common_samples, relevant_metadata$sample_id), ]
    
    relevant_metadata$plate_id   <- as.factor(relevant_metadata$plate_id)
    relevant_metadata$dose_level <- as.factor(relevant_metadata$dose_level)
    
    stopifnot(all(relevant_metadata$sample_id == colnames(count_matrix)))
    
    samples <- unique(relevant_metadata$epa_sample_id)
    samples <- samples[!(samples %in% c("DMSO", "", NA))]
    
    sample_results <-  list()
    
    for (i in seq_along(samples)) {
        sample <- samples[i]
        cat("\n--> ", sample, " (", i, "/", length(samples), ")\n", sep = "")
        
        sub_relevant_metadata <- relevant_metadata[relevant_metadata$epa_sample_id == sample | (relevant_metadata$stype == "vehicle control"), ]
        sub_relevant_count_data <- count_matrix[, sub_relevant_metadata$sample_id, drop = FALSE]
        
        dds <- DESeqDataSetFromMatrix(countData = sub_relevant_count_data,
                                      colData   = sub_relevant_metadata,
                                      design    = ~ plate_id + dose_level)
        
        keep_genes <- rowMeans(counts(dds)) >= 5 # as in paper
        dds <- dds[keep_genes, ]
        
        dds <- DESeq(dds, betaPrior = FALSE, parallel = TRUE)
        
        current_levels <- unique(sub_relevant_metadata$dose_level)
        dose_levels <- setdiff(current_levels, "0")
        
        res_list <- bplapply(dose_levels, function(dose) {
            
            res <- suppressMessages(lfcShrink(dds, contrast = c("dose_level", dose, "0"), type = "ashr"))
            res_df <- as.data.frame(res)
            res_df$dose_level <- dose
            res_df <- res_df %>%
                rownames_to_column("probe_id") %>%
                dplyr::left_join(manifest[, c("Probe_Name", "Gene_Symbol", "Entrez_ID", "Probe_Flag")],   by = c("probe_id" = "Probe_Name")) %>%
                dplyr::filter(Probe_Flag != "bad")
            res_df
        })
        
        res_combined <- do.call(rbind, res_list)
        res_combined$sample_id <- sample
        res_combined$pg_id <- pg
        
        sample_results[[sample]] <- res_combined
        
    }
    
    all_results[[pg]] <- do.call(rbind, sample_results)
    cat("\n")
}
```




# AGGREGATING PROBES TO GENE LEVEL

In the paper, there was mentioned: "DESeq2 probe-level L2FC results for each test chemical were summarized as a matrix in which the rows and columns correspond to the treatment concentrations and probes, respectively. For each probe-level L2FC matrix, L2FC values were aggregated to the gene level by using the highest magnitude fold change for any associated probe in either direction."
Here we will do the same to get gene-level results instead of probe-level.

```{r}
process_sample <- function(sample, plate, plategroup) {
    library(gtools)
    library(dplyr)
    
    sample_plate <- plate[plate$sample_id == sample, ]
    dose_levels <- mixedsort(unique(sample_plate$dose_level))
    
    sample_results <- list()
    
    for (dose in dose_levels) {
        dose_plate <- sample_plate[sample_plate$dose_level == dose, ]
        
        res_df <- dose_plate %>%
            dplyr::select(Gene_Symbol, 
                          padj, 
                          log2FoldChange) %>%
            dplyr::mutate(abs_L2FC = abs(log2FoldChange)) %>%
            dplyr::group_by(Gene_Symbol) %>%
            dplyr::slice_max(order_by = abs_L2FC, 
                             with_ties = FALSE) %>%
            dplyr::ungroup() %>%
            dplyr::mutate(sample_id = sample,
                          plate_group = plategroup,
                          dose_level = dose,
                          sample_dose = paste0(sample, "_", dose),
                          gene_symbol = Gene_Symbol, 
                          log2foldchange = log2FoldChange) %>%
            dplyr::select(sample_dose, 
                          gene_symbol, 
                          log2foldchange, 
                          padj, 
                          plate_group, 
                          sample_id, 
                          dose_level)
        
        sample_results[[dose]] <- res_df
    }
    
    return(sample_results)
}
```

```{r}
aggregated_results_all <- list()
plategroups <- seq_along(all_results)

for (plategroup in plategroups) {
    cat("Processing plate group:", plategroup, "\n")
    
    plate <- as.data.frame(all_results[[plategroup]])
    samples <- unique(plate$sample_id)
    
    bp_param <- MulticoreParam(workers = 40)
    
    sample_results_list <- bplapply(samples, function(sample) {
        process_sample(sample, plate, plategroup)
    }, BPPARAM = bp_param)
    
    for (sample_result in sample_results_list) {
        aggregated_results_all <- c(aggregated_results_all, sample_result)
    }
}
```

```{r}
combined_long_df <- bind_rows(aggregated_results_all)

log2_wide_paired <- as.data.frame(combined_long_df %>%
                                      group_by(sample_dose) %>%
                                      select(sample_dose, gene_symbol, log2foldchange) %>%
                                      pivot_wider(names_from = gene_symbol, values_from = log2foldchange, names_prefix = "lfc_"))

padj_wide_paired <- as.data.frame(combined_long_df %>%
                                      select(sample_dose, gene_symbol, padj) %>%
                                      pivot_wider(names_from = gene_symbol, values_from = padj, names_prefix = "padj_"))

final_wide <- log2_wide_paired %>% 
    left_join(padj_wide_paired, by = "sample_dose")

final_wide
```




# SAVE 

```{r}
cell_name <- "MCF7"
output_file <- paste0("../output/aggregated_results_", cell_name, ".feather")

write_feather(final_wide, output_file)
```
